Quantum H theorem, variational entropy, and relaxation processes

### 2021-01-11 tex/quantum.tex
Rewording: up to quantum H-theorem proof.

We propose the following functional as an alternative $H$-functional
for quantum non-homogeneous diluted gases:
%
\begin{eqnarray}\label{eq:qHdef}
    \mathcal{H} (t)&=&\sum_{M=1}^{K} \sum_{n} \bigg[ f_{Mn}(t) \ln f_{Mn}(t)\bigg.\nonumber \\
    &&\bigg.\pm \Big(1 \mp f_{Mn}(t)) \ln (1 \mp f_{Mn}(t)\Big) \Big]   \delta V_M.
\end{eqnarray}
%
Here, the upper and lower sign refer to \commr{bosons and fermions respectively}.
In addition, when needed, each cell will have an associated local chemical potential,
$\alpha_M$, and its local $H$-functional, which is defined by:
%
\begin{equation}\label{eq:qHMdef}
   \mathcal{H}_M(t)=\sum_{n} \bigg[ f_{Mn}(t) \ln f_{Mn}(t)
   \pm \Big(1 \mp f_{Mn}(t)\Big) \ln \Big(1 \mp f_{Mn}(t)\Big) \bigg] \de V_M.
\end{equation}
%
Therefore, $\mathcal{N}_M$ and $\mathcal{E}_M$ as functions of time are given by:
%
\begin{subequations}
\begin{equation}
    {\mathcal{N}}_M(t) = \sum_{n}f_{Mn}(t) \delta V_M
\end{equation}
%
and
%
\begin{equation}
\mathcal{E}_M(t) = \sum_{n}f_{Mn}(t)\epsilon_{n} \de V_M
\end{equation}
\end{subequations}
%
which are, in addtion, constrained by the global microcanonical restrictions
%
\begin{subequations}
\begin{equation}\label{eq:qGlobRestN}
    \sum_{M=1}^{K}\left[\sum_{n}f_{Mn}(t)\right]\delta  V_M
    =\sum_{M=1}^{K}\mathcal{N}_{M}(t)\delta V_M=N
\end{equation}
%
and
%
\begin{equation}\label{eq:qGlobRestE}
    \sum_{M=1}^{K}\left[\sum_{n}f_{Mn}(t)\epsilon_{n}\right]\delta V_M
    =\sum_{M=1}^{K}\mathcal{E}_M(t)\delta V_M=E. 
\end{equation}
\end{subequations}
%

@@

%============================================================
\subsubsection{Properties of $\mathcal{H}$ in equilibrium}
%============================================================
If the system is in equilibrium (at $t\to\infty$), the temperature becomes homogeneous throughout
the complete system. Also, the local number of particles, the local
energy, and the Lagrange multipliers do not depend on the cell number and they should be homogeneous,
this is \commr{[This needs a rationale\dots]}
%
\begin{subequations}\label{eq:qEqRestrictions}
\begin{equation}
   {\mathcal{N}}_M(t\to\infty)\equiv \bar{\mathcal{N}}=N/K,
\end{equation}
%
%
\begin{equation}
	{\mathcal{E}}_M(t\to\infty)\equiv \bar{\mathcal{E}}=E/K,
\end{equation}
\end{subequations}
%
%
\begin{subequations}\label{eq:qEqAlphaBeta}
\begin{equation}
	\alpha_M=\bar\alpha,\quad\forall M,
\end{equation}
%
and
\begin{equation}
	\beta_M=\bar\beta, \quad\forall M.
\end{equation}
\end{subequations}
%
Substituting Eqs.~(\ref{eq:qEqAlphaBeta}) into Eq.~(\ref{eq:qfMn}) yields the distribution
function of each cell in equilibrium to be
%
\begin{equation}\label{eq:qfneq}
    \bar f_{Mn}=\bar f_n =\frac{1}{\exp\big(-\bar\alpha-\bar\beta \epsilon_n\big)\pm 1},\quad\forall M.
\end{equation}
%
By means of the above equation, we can recover the distribution function and the entropy of a diluted quantum gas in equilibrium as follows. Setting $\bar\alpha=\mu/kT$ and $\bar\beta=-1/kT$,
and substituting them in Eq.~(\ref{eq:qfneq}) renders the Fermi-Dirac and Bose-Einstein
distribution functions:
%
\begin{equation}
    \bar{f}_{n}=\frac{1}{\exp\big(\frac{{\epsilon_n}-\mu}{kT}\big)\pm 1},
\end{equation}
%
and substituting Eq.~(\ref{eq:qfneq}) into the negative of Eq.(\ref{eq:qHdef}) provides the entropy:
%
\begin{eqnarray}\label{eq:varEntropyDef}
      S&=&\sum_{M=1}^{K}
        \sum_n  \left[
        		\left(\frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)
           \commr{\ln \left(\frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)}
         \right]\nonumber \\
      &&\quad\pm  \ln \left[\prod_{M=1}^{K} \prod_{n}
         \left(1 \mp \frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)
         \right] \delta V_M.
\end{eqnarray}
%
For obtaining the last result, we have used the well-known fact that the $H$-functional
is the negative of the entropy. The property provided by Eq.(\ref{eq:varEntropyDef})
is precisely what we call the ``variational entropy'',
and this name reflects the fact that it was obtained \textit{via}
the variational method.

@@

@@

We rewrite the \textit{variational entropy} (\ref{eq:qHdef}) as
%
\begin{equation}
    \mathcal{H}(t)=\sum_{M=1}^{K} \sum_{n} \left[ f_{Mn} \ln f_{Mn} \pm (1 \mp f_{Mn}) \ln (1 \mp f_{Mn}) \right] \de V_M \delta \epsilon_n \label{entropy2},
\end{equation}
%
where, for simplicity, we have defined $f(\epsilon_{n},t)\equiv f_{Mn}$.

This distribution function must satisfy the local equilibrium hypothesis in
each cell ($f_{nM}$ must be spatial-homogeneous for each $M$), the
non-homogeneous distribution function assumption in the total volume ($f_{nM}$
must be different among all cells), and the following restrictions
%
\begin{eqnarray}
        \sum_{n}f_{Mn} \delta \epsilon_n=\bar{\mathcal{N}}+\Delta_M(t); \ \ \ \ \sum_{n}\epsilon_{n}f_{Mn} \delta \epsilon_n=\bar{\mathcal{E}}+ \delta_M(t), \label{restrictionoutside}
  \end{eqnarray}
%
where $\bar {\mathcal{N}}$ and $\bar{\mathcal{E}}$ are the local particle
number and the local energy in equilibrium
%
\begin{equation}
      \bar{\mathcal{N}}= \sum_n \bar{f}_n \delta \epsilon_n; \ \ \ \ \bar{\mathcal{E}}= \sum_n \epsilon_n\bar{f}_n \delta \epsilon_n,
  \end{equation}
%
and $\Delta_M,\delta_M$ could be seen as a deviation from $\bar{\mathcal{N}}$
and $\bar{\mathcal{E}}$ respectively, with $\Delta_M(t)\ll \bar{\mathcal{N}}$
and $\delta_M(t) \ll \bar{\mathcal{E}}$. The previous conditions and
restrictions are the quantum analogous case of the classical case. Also, this
system will suffer, as in the classical case, a \textit{relaxation process}.

  The quantities $\Delta_M,\delta_M$ are sufficiently big to be different from
fluctuations in the system but sufficiently small to consider the system not
far from the equilibrium state, as in the classical case.

If we perform the derivative of the variational entropy with respect to time,
we obtain
%
\begin{equation}
   \frac{d \mathcal{H} (t)}{dt}= \sum_n \sum_{M=1}^{K} \dot{f}_{nM}(t)\ln \left[ \frac{f_{nM}(t)}{1\mp f_{nM}(t)} \right] \de V_M \delta \epsilon_n.\label{deltaH}
\end{equation}
%
Using the first-order approximation (\ref{firstorder}), we get
%
\begin{eqnarray}
    \frac{d\mathcal{H} (t)}{dt}&=&\sum_n \sum_{M=1}^{K} \bar{f}_{n}\ln \left[ \frac{\bar{f}_{n}(1+g_{nM})}{1\mp \bar{f}_{n} (1+ g_{nM})} \right]\dot{g}_{nM} \de V_M\delta \epsilon_n \nonumber \\
    &=&\sum_n \sum_{M=1}^{K} \bar{f}_n \left \{ \ln [\bar{f}_n+\bar{f}_n g_{nM}]\dot{g}_{nM}-\ln [1\mp\bar{f}_n\mp\bar{f}_n g_{nM}]\dot{g}_{nM}  \right \}\de V_M \delta \epsilon_n.\nonumber \\
    \label{cambioH1}
\end{eqnarray}
%
It's necessary to remark that, in the case of fermions
%
\begin{eqnarray}
   1-\bar f_n -\bar f_n g_{nM}>0 \ \ \Rightarrow \ \ \frac{1}{\bar f_n}>1+g_{nM}. \label{fermionrestriction}
\end{eqnarray}
%
The previous expression establishes that the value of $g_{nM}$ is determined by
$\bar f_n$. Furthermore, whether $\bar f_{n}=1$, this is, all lower energy
levels are occupied, then the system will not have inhomogeneities because of
the exclusion principle, and as a consequence of this,  $g_{nM}=0$. This result
corresponds to the system in the zero temperature condition. We can observe
that in this condition, (\ref{fermionrestriction}) is violated. Due to this,
the zero temperature condition is excluded.

We can observe in the case of bosons, if $\bar{f}_n \gg \bar{f}_n |g_{nM}|$
then $1+\bar{f}_n \gg \bar{f}_n |g_{nM}|$. In fermions, in the case of a
non-extremely degenerated condition, it is true that $1-\bar{f}_n \gg \bar{f}_n
|g_{nM}|$. With those relations, we can approximate the logarithm functions in
(\ref{cambioH1}) to their first-order Taylor series around $\bar f_n g_{nM}=0$
%
\begin{equation}
    \ln [\bar{f}_n+\bar{f}_n g_{nM}] \approx \ln [\bar{f}_n]+ g_{nM}; \ \ \ \ \ln[1\mp\bar{f}_n\mp\bar{f}_n g_{nM}] \approx \ln[1\mp\bar{f}_n]\mp\frac{\bar{f}_n}{1\mp\bar{f}_{n}} g_{nM}. \label{lnapproximation}
\end{equation}
%
With the previous approximation, (\ref{cambioH1}) will be
%
\begin{eqnarray}
    \frac{d\mathcal{H}}{dt}&=&\sum_n \sum_{M=1}^{K} \bar{f}_n\left \{ (\ln \bar{f}_n+ g_{nM})\dot{g}_{nM}\right\} \de V_M \delta \epsilon_n \nonumber \\
    &&-\sum_{n}\sum_{M=1}^{K}\bar f_n\left\{ \left( \ln[1\mp\bar{f}_n]\mp \left[\frac{\bar{f}_n}{1\mp\bar{f}_n} \right] g_{nM}\right)\dot{g}_{nM} \right \}\de V_M \delta \epsilon_n.\label{cambioH2}
\end{eqnarray}
%
Making use of the expression (\ref{relation}),
(\ref{cambioH2}) casts into
%
\begin{eqnarray}
    \frac{d\mathcal{H}}{dt}&=&\sum_n \sum_{M=1}^{K} \bar{f}_n\left \{ (\bar{\alpha}+\bar{\beta}{\epsilon}_n)\dot{g}_{nM}+ g_{nM}\left(1\pm e^{\bar{\alpha}+\bar{\beta}{\epsilon}_n}\right)\dot{g}_{nM} \right \} \de V_M \delta \epsilon_n. \nonumber \\
    \label{cambioH3}
\end{eqnarray}
%
On the other hand, we obtain from the restrictions the following expressions
%
\begin{eqnarray}
    &&\sum_n \bar{f}_n g_{nM} \delta \epsilon_n=\Delta_M(t) \ \  \Rightarrow \ \  \sum_n \bar{f}_n \dot{g}_{nM} \delta \epsilon_n=\dot{\Delta}_M(t), \nonumber \\
    &&\sum_n  \bar{f}_n g_{nM}\epsilon_n \delta \epsilon_n=\delta_M(t) \ \  \Rightarrow \ \  \sum_n \bar{f}_n \dot{g}_{nM}\epsilon_n \delta \epsilon_n=\dot{\delta}_M(t)
\end{eqnarray}
%
and as a consequence of $\sum_{M=1}^{K} \Delta_M(t) \de V_M  =\sum_{M=1}^{K} \delta_M(t) \de V_M =0$, we find
%
\begin{equation}
    \sum_{M=1}^{K} \dot{\Delta}_M(t) \de V_M =\sum_{M=1}^{K} \dot{\delta}_{M}(t) \de V_M=0.
\end{equation}
%
Substituting the previous expression to (\ref{cambioH3}) we obtain
%
\begin{eqnarray}
   \frac{d\mathcal{H}}{dt}&=&  \sum_n e^{\bar{\alpha}+\bar{\beta}\epsilon_n}\sum_M  g_{nM}\dot{g}_{nM} \de V_M\delta \epsilon_n. \nonumber \\ \label{cambioH4}
\end{eqnarray}
%
Now, the summation over $M$ will be expressed in two summations, such as in the
classical case
%
\begin{equation}
    \frac{d\mathcal{H}}{dt}=\sum_n  e^{\bar{\alpha}+\bar{\beta}\epsilon_n}\left(\sum_J ^{L} g^{+}_{nJ}\dot{g}^{+}_{nJ}\de V_J \delta \epsilon_n+\sum^{P}_J  g^{-}_{nJ}\dot{g}^{-}_{nJ} \de V_J\delta \epsilon_n \right), \label{cambioH5}
\end{equation}
%
where $L+P=K$. Besides, $\dot{g}^{+}_{J}$ represents the change on the
deviation on cells that have an excess of particles or energy while
$\dot{g}^{-}_{J}$  represents the change on the deviations on cells that have
missing particles or energy.
On the other hand, $g^{+}_{J}$  represents the value of the deviation on cells
that have an excess of particles or energy. In contrast, $g^{-}_{J}$ represents
the value of the deviation on cells that have missing particles or energy.

@@

When the system is in equilibrium, $g_{nM}=0$ and in consequence of that, $\dot
g_{nM}=0$ and finally from (\ref{cambioH1}) $\frac{d\mathcal{H}}{dt}=0$. Then, we can
say that any ideal quantum gas perturbed always evolves to the equilibrium
state, such as the classical $H$-theorem. Then, we obtained the quantum version
of the Boltzmann's $H$-theorem including inhomogeneities in the system.

>>>>

We propose the following functional as an alternative $H$-functional
for quantum non-homogeneous diluted gases:
%
\begin{eqnarray}\label{eq:qHdef}
    \mathcal{H} (t)&=&\sum_{M=1}^{K} \sum_{n} \bigg[ f_{Mn}(t) \ln f_{Mn}(t)\bigg.\nonumber \\
    &&\qquad\qquad\bigg.\pm \Big(1 \mp f_{Mn}(t)) \ln (1 \mp f_{Mn}(t)\Big) \Big]   \delta V_M.
\end{eqnarray}
%
Here, the upper and lower sign refer to \commr{bosons and fermions respectively}.
In addition, when needed, each cell will have an associated local chemical potential,
$\alpha_M$, and a local $H$-functional, which is defined by:
%
\begin{equation}\label{eq:qHMdef}
   \mathcal{H}_M(t)=\sum_{n} \bigg[ f_{Mn}(t) \ln f_{Mn}(t)
   \pm \Big(1 \mp f_{Mn}(t)\Big) \ln \Big(1 \mp f_{Mn}(t)\Big) \bigg] \de V_M.
\end{equation}
%
Therefore, $\mathcal{N}_M$ and $\mathcal{E}_M$ as functions of time are given by:
%
\begin{subequations}
\begin{equation}
    {\mathcal{N}}_M(t) = \sum_{n}f_{Mn}(t) \delta V_M
\end{equation}
%
and
%
\begin{equation}
\mathcal{E}_M(t) = \sum_{n}f_{Mn}(t)\epsilon_{n} \de V_M
\end{equation}
\end{subequations}
%
which are, in addtion, constrained by the global microcanonical restrictions
%
\begin{subequations}\label{eq:qGlobRest}
\begin{equation}\label{eq:qGlobRestN}
    \sum_{M=1}^{K}\left[\sum_{n}f_{Mn}(t)\right]\delta  V_M
    =\sum_{M=1}^{K}\mathcal{N}_{M}(t)\delta V_M=N
\end{equation}
%
and
%
\begin{equation}\label{eq:qGlobRestE}
    \sum_{M=1}^{K}\left[\sum_{n}f_{Mn}(t)\epsilon_{n}\right]\delta V_M
    =\sum_{M=1}^{K}\mathcal{E}_M(t)\delta V_M=E. 
\end{equation}
\end{subequations}
%

@@

%============================================================
\subsubsection{Properties of $\mathcal{H}$ for systems in equilibrium}
%============================================================
If the system is in equilibrium (at $t\to\infty$), the temperature becomes homogeneous throughout
the complete system. Also, the local number of particles, the local
energy, and the Lagrange multipliers do not depend on the cell number and they should be homogeneous,
this is \commr{[This needs a rationale\dots]}
%
\begin{subequations}\label{eq:qEqRestrictions}
\begin{equation}
   {\mathcal{N}}_M(t\to\infty)\equiv \bar{\mathcal{N}}=N/K,
\end{equation}
%
%
\begin{equation}
	{\mathcal{E}}_M(t\to\infty)\equiv \bar{\mathcal{E}}=E/K,
\end{equation}
\end{subequations}
%
%
\begin{subequations}\label{eq:qEqAlphaBeta}
\begin{equation}
	\alpha_M=\bar\alpha,\quad\forall M,
\end{equation}
%
and
\begin{equation}
	\beta_M=\bar\beta, \quad\forall M.
\end{equation}
\end{subequations}
%
Substituting Eqs.~(\ref{eq:qEqAlphaBeta}) into Eq.~(\ref{eq:qfMn}) yields the distribution
function of each cell in equilibrium to be
%
\begin{equation}\label{eq:qfneq}
    \bar f_{Mn}=\bar f_n =\frac{1}{\exp\big(-\bar\alpha-\bar\beta \epsilon_n\big)\pm 1},\quad\forall M.
\end{equation}
%
By means of the above equation, we can recover the distribution function and the entropy of a diluted quantum gas in equilibrium as follows. Setting $\bar\alpha=\mu/kT$ and $\bar\beta=-1/kT$,
and substituting them in Eq.~(\ref{eq:qfneq}) renders the Fermi-Dirac and Bose-Einstein
distribution functions:
%
\begin{equation}
    \bar{f}_{n}=\frac{1}{\exp\big(\frac{{\epsilon_n}-\mu}{kT}\big)\pm 1},
\end{equation}
%
and substituting Eq.~(\ref{eq:qfneq}) into the negative of Eq.(\ref{eq:qHdef}) provides the entropy:
%
\begin{eqnarray}\label{eq:varEntropyDef}
      S&=&\sum_{M=1}^{K}
        \sum_n  \left[
        		\left(\frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)
           \commr{\ln \left(\frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)}
         \right]\nonumber \\
      &&\quad\pm  \ln \left[\prod_{M=1}^{K} \prod_{n}
         \left(1 \mp \frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)
         \right] \delta V_M.
\end{eqnarray}
%
For obtaining the last result, we have used the well-known fact that the $H$-functional
is the negative of the entropy. The property provided by Eq.(\ref{eq:varEntropyDef})
is precisely what we refer to as ``variational entropy'',
and this name reflects the fact that it was obtained \textit{via}
the variational method.

@@

For quantum systems, we also accept the validity of the local equilibrium hypothesis
for every cell in the system. This allows us to define non-homogeneous systems,
wherein thermodynamic quantities are well-defined on a cell-per-cell basis,
and in terms of the equilibrium properties, we have

%
\begin{subequations}\label{eq:restrictionoutside}
\begin{equation}
        \mathcal{N}_M(t)=\sum_{n}f_{Mn}(t)=\bar{\mathcal{N}}+\Delta_M(t) 
\end{equation}
  and
\begin{equation}
        \mathcal{E}_M(t)=\sum_{n}\epsilon_{n}f_{Mn}(t)=\bar{\mathcal{E}}+ \delta_M(t).
\end{equation}
\end{subequations}
%
In Eqs.~(\ref{eq:restrictionoutside}) $\bar {\mathcal{N}}$ and $\bar{\mathcal{E}}$ are the local particle
number and the local energy of cells in equilibrium, which are given by Eqs.~(\ref{eq:qEqRestrictions}),
and $\Delta_M$ and $\delta_M$ are deviations from $\bar{\mathcal{N}}$
and $\bar{\mathcal{E}}$, respectively, with $\Delta_M(t)\ll \bar{\mathcal{N}}$
and $\delta_M(t) \ll \bar{\mathcal{E}}$.

In the present context, $|\Delta_M|$ and $|\delta_M|$ are sufficiently large so they are not
fluctuations of the system, but sufficiently small so that the local equilibrium hypothesis is
valid for $t>0$ (we set $t_0=0$, and $t_0$ is the initial time at which the system is prepared).
Therefore, it is reasonable to re-write the distribution functions as:
%
\begin{equation}\label{eq:qFirstOrd}
   f_{Mn}(t)=\bar{f}_n(1+g_{Mn}(t)),\quad
   1\gg|g_{Mn}(t)|,
\end{equation}
%
from which it follows, by substituting Eq.~(\ref{eq:qFirstOrd}) into
Eq.~(\ref{eq:restrictionoutside}), that $\Delta_M$ and $\delta_M$ satisfy:
%
\begin{subequations}
\begin{equation}
    \Delta_M(t)=\sum_n \bar{f}_n g_{nM}
\end{equation}
%
and
%
\begin{equation}
	\delta_M(t)=\sum_n  \bar{f}_n g_{nM}\epsilon_n.
\end{equation}
\end{subequations}
%

\commr{
An additional consideration is necessary for treating gases composed of fermions,
which must obey the relation $f_{Mn}(t)\leq1$, hence
%
\begin{eqnarray}\label{eq:fermionrestriction}
   1-\bar f_n -\bar f_n g_{nM}\geq0 \ \ \Rightarrow \ \ \frac{1}{\bar f_n}\geq1+g_{nM}.
\end{eqnarray}
%
$\bar f_{n}=1$ is certainly satisfied if the system is at zero temperature. In this state,
all energy levels below and including the Fermi energy are occupied because of
the exclusion principle, therefore the system will
necessarily be homogeneous , and consequently $g_{nM}=0$. In this article, we will
omit this scenario and will only discuss Fermi gases whose temperature is non zero.
}

To proof the quantum $H$-theorem, start by taking the time-derivative of Eq. (\ref{eq:qHdef}):
%
\begin{equation}\label{eq:deltaH}
   \frac{d \mathcal{H} (t)}{dt}= \sum_n \sum_{M=1}^{K} \dot{f}_{nM}(t)\ln \left[ \frac{f_{nM}(t)}{1\mp f_{nM}(t)} \right] \de V_M.
\end{equation}
%
Subsequently, substitute Eq.~(\ref{eq:qFirstOrd}) in the above equation to obtain:
%
\begin{eqnarray}\label{eq:cambioH1}
    \frac{d\mathcal{H} (t)}{dt}&=&
      \sum_n \sum_{M=1}^{K} \bar{f}_{n}\ln \left[
        \frac{\bar{f}_{n}(1+g_{nM})}{1\mp \bar{f}_{n} (1+ g_{nM})}
      \right]\dot{g}_{nM} \delta V_M \nonumber \\
    &=&\sum_n \sum_{M=1}^{K} \bar{f}_n \left \{
      \ln [\bar{f}_n+\bar{f}_n g_{nM}]\dot{g}_{nM}
      -\ln [1\mp\bar{f}_n\mp\bar{f}_n g_{nM}]\dot{g}_{nM}
    \right \}\delta V_M,
\end{eqnarray}
%
Approximate the logarithmic terms, corresponding to
Fermi and Bose gases, through a Taylor series around $\bar f_n g_{nM}=0$, and
substitute this into Eq.~(\ref{eq:cambioH1}), which gives:
%
\begin{subequations}\label{eq:qlnApprox}
\begin{equation}\label{eq:qlnApproxFermions}
	\ln[1\mp\bar{f}_n\mp\bar{f}_n g_{nM}]
    	\approx \ln[1\mp\bar{f}_n]\mp\frac{\bar{f}_n}{1\mp\bar{f}_{n}} g_{nM}
\end{equation}
%
and
%
\begin{equation}\label{eq:qlnApproxBosons}
    \ln [\bar{f}_n+\bar{f}_n g_{nM}] \approx \ln [\bar{f}_n]+ g_{nM},
\end{equation}
\end{subequations}
%
respectively. Eq.~(\ref{eq:qlnApproxFermions}) is valid because, for non-extremely degenerated Fermi gases, 
$1-\bar{f}_n \gg \bar{f}_n|g_{nM}|$, and Eq.~(\ref{eq:qlnApproxBosons}) is fulfilled because,
for Boson gases,
$1+\bar{f}_n \gg \bar{f}_n |g_{nM}|$ when $\bar{f}_n \gg \bar{f}_n |g_{nM}|$.

Combine Eqs. (\ref{eq:cambioH1}) and (\ref{eq:qlnApprox}):
%
\begin{eqnarray}\label{eq:cambioH2}
    \frac{d\mathcal{H}}{dt}&=&\sum_n \sum_{M=1}^{K} \bar{f}_n\left \{ (\ln \bar{f}_n+ g_{nM})\dot{g}_{nM}\right\} \de V_M \delta \epsilon_n \nonumber \\
    &&-\sum_{n}\sum_{M=1}^{K}\bar f_n\left\{ \left( \ln[1\mp\bar{f}_n]\mp \left[\frac{\bar{f}_n}{1\mp\bar{f}_n} \right] g_{nM}\right)\dot{g}_{nM} \right \}\de V_M,
\end{eqnarray}
%
and substitute Eq.~(\ref{eq:qfneq}) into Eq.~(\ref{eq:cambioH2}):
%
\begin{equation}\label{eq:cambioH3}
    \frac{d\mathcal{H}}{dt}=
       \sum_n \sum_{M=1}^{K} \bar{f}_n\left \{
          (\bar{\alpha}+\bar{\beta}{\epsilon}_n)\dot{g}_{nM}
           +g_{nM}\left(1\pm e^{\bar{\alpha}+\bar{\beta}{\epsilon}_n}\right)\dot{g}_{nM}
       \right \} \de V_M.
\end{equation}
%

Since both the total number of particles and the total energy of the system are constant, it follows
from Eqs.~(\ref{eq:qGlobRest}) and (\ref{eq:restrictionoutside}) that:
%
\begin{subequations}
\begin{equation}
\frac{dN}{dt}=\sum_{M=1}^K\dot{\mathcal{N}}_M\delta V_M
   =\sum_{M=1}^K\sum_n \bar{f}_n \dot{g}_{nM}\delta V_M=\sum_{M=1}^K\dot{\Delta}_M(t)\delta V_M=0
\end{equation}
%
and
%
\begin{equation}
\frac{dE}{dt}=\sum_{M=1}^K\dot{\mathcal{E}}_M\delta V_M
   =\sum_{M=1}^K\sum_n \bar{f}_n \dot{g}_{nM}\epsilon_n\delta V_M=\sum_{M=1}^K\dot{\delta}_M(t)\delta V_M=0.
\end{equation}
\end{subequations}
%

Substitute the previous expression in Eq.~(\ref{eq:cambioH3}) to obtain:
%
\begin{equation}\label{eq:cambioH4}
   \frac{d\mathcal{H}}{dt}=\sum_n e^{\bar{\alpha}+\bar{\beta}\epsilon_n}
   \sum_M  g_{nM}\dot{g}_{nM} \delta V_M \leq 0.
\end{equation}
%

For obtaining the right-most hand side of Eq.~(\ref{eq:cambioH4}), 
we have exploited the relation $g_{nM}\dot{g}_{nM}\leq0$ for $t>0$. This can be proven 
by simply arguing that, in the initial
state, if a cell is described by $g_{nM}(t_0)>0$ then $g_{nM}(t)\geq0$ and $\dot g_{nM}(t)\leq0$, and if
$g_{nM}(t_0)<0$ then $g_{nM}(t)\leq0$ and $\dot g_{nM}(t)\geq0$. Here we have used the facts that
the system in equilibrium is homogeneous, and that, by accepting the local equilibrium hypothesis, $g_{Mn}(t)$
is a monotonic function and $g_{nM}\to0$ as $t\to\infty$ as the system aproaches to the equilibrium state.
Another approach to prove Eq.~(\ref{eq:cambioH4}) consists of splitting the cells into two subsets,
such as we did in the classical scenario, however, we will ommit this approach in this section.

@@

%
Now, the summation over $M$ will be expressed in two summations, such as in the
classical case
%
\begin{equation}
    \frac{d\mathcal{H}}{dt}=\sum_n  e^{\bar{\alpha}+\bar{\beta}\epsilon_n}\left(\sum_J ^{L} g^{+}_{nJ}\dot{g}^{+}_{nJ}\de V_J \delta \epsilon_n+\sum^{P}_J  g^{-}_{nJ}\dot{g}^{-}_{nJ} \de V_J\delta \epsilon_n \right), \label{cambioH5}
\end{equation}
%
where $L+P=K$. Besides, $\dot{g}^{+}_{J}$ represents the change on the
deviation on cells that have an excess of particles or energy while
$\dot{g}^{-}_{J}$  represents the change on the deviations on cells that have
missing particles or energy.
On the other hand, $g^{+}_{J}$  represents the value of the deviation on cells
that have an excess of particles or energy. In contrast, $g^{-}_{J}$ represents
the value of the deviation on cells that have missing particles or energy.

@@

When the system is in equilibrium, $g_{nM}=0$ and in consequence of that, $\dot
g_{nM}=0$ and finally from (\ref{eq:cambioH1}) $\frac{d\mathcal{H}}{dt}=0$. Then, we can
say that any ideal quantum gas perturbed always evolves to the equilibrium
state, such as the classical $H$-theorem. Then, we obtained the quantum version
of the Boltzmann's $H$-theorem including inhomogeneities in the system.
